---
title: Choosing a Basis Family
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
resource_files:
- albers.css
- albers.js
- albers-header.html
params:
  family: red
  preset: homage

vignette: |
  %\VignetteIndexEntry{Choosing a Basis Family} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r library}
library(fmrilatent)
```

Basis choice is a modeling decision. You are choosing which structure is cheap
to represent: smooth time courses, band-limited time courses, local parcel
patterns, or smooth spatial atoms. This vignette uses one toy dataset to show
how to compare candidates before moving to larger fMRI data.

## What signal are we trying to represent?

The dataset mixes two slow temporal signals with parcel-specific spatial
weights and a small amount of noise.

```{r make-signal}
mask <- array(TRUE, dim = c(4, 4, 4))
mask_vol <- neuroim2::LogicalNeuroVol(mask, neuroim2::NeuroSpace(dim(mask)))

n_time <- 32L
n_vox <- sum(mask)
parcel_map <- as.integer(rep_len(1:4, n_vox))

set.seed(11)
t <- seq_len(n_time)
signal <- cbind(sin(2 * pi * t / n_time), cos(4 * pi * t / n_time))
weights <- matrix(rnorm(2 * n_vox), nrow = 2)
X <- signal %*% weights + matrix(rnorm(n_time * n_vox, sd = 0.05), nrow = n_time)
```

## Which temporal basis fits this shape?

Temporal bases share one set of time functions across all voxels. Here we keep
the comparison small and report reconstruction error directly.

```{r temporal-compare}
fits <- list(
  dct = encode(X, spec_time_dct(k = 6), mask = mask_vol, materialize = "matrix"),
  slepian = encode(X, spec_time_slepian(tr = 2, bandwidth = 0.08, k = 4),
                   mask = mask_vol, materialize = "matrix"),
  bspline = encode(X, spec_time_bspline(k = 6), mask = mask_vol,
                   materialize = "matrix")
)

rmse <- function(x) sqrt(mean((as.matrix(x) - X)^2))
data.frame(
  family = names(fits),
  components = vapply(fits, function(x) ncol(basis(x)), integer(1)),
  rmse = round(vapply(fits, rmse, numeric(1)), 4)
)
```

```{r validate-temporal, include = FALSE}
temporal_errors <- vapply(fits, rmse, numeric(1))
stopifnot(all(is.finite(temporal_errors)), all(temporal_errors >= 0))
```

```{r temporal-plot, echo = FALSE, fig.width = 7, fig.height = 3.4, fig.alt = "One voxel: original signal versus DCT, Slepian, and B-spline reconstructions."}
op <- par(mar = c(4, 4, 1, 1))
cols <- c(dct = "firebrick", slepian = "steelblue", bspline = "darkgreen")
plot(X[, 1], type = "l", col = "grey60", lwd = 3,
     xlab = "time point", ylab = "signal")
for (nm in names(fits)) lines(as.matrix(fits[[nm]])[, 1], col = cols[[nm]], lwd = 2)
legend("topright", c("original", names(fits)),
       col = c("grey60", cols), lwd = c(3, 2, 2, 2), bty = "n", cex = 0.85)
par(op)
```

On this toy signal DCT wins numerically because the signal *is* two clean
cosines — exactly what a cosine basis represents with few components. That is
a property of the example, not a universal ranking. The recommendation below
reflects the priors that hold on real fMRI data, where the signal rarely
matches one basis so cleanly:

| Signal property | Prefer | Why | Avoid when |
|---|---|---|---|
| Energy in a known frequency band | Slepian/DPSS | Optimal bandlimiting, minimal spectral leakage | Signal is broadband |
| No strong spectral prior | DCT | Fast, orthogonal, parameter-free beyond `k` | You need localized, non-periodic structure |
| Smooth drifts / hemodynamic shapes | B-spline | Smooth, localized, non-periodic | Sharp or oscillatory features dominate |

Use DCT as a simple default, Slepian when frequency concentration matters, and
B-splines when smooth local variation is the main prior.

## What changes for spatial bases?

Spatial bases share one set of voxel patterns across time. Parcel PCA is
data-driven and local to a reduction; HRBF produces smooth spatial atoms from
the mask geometry.

```{r spatial-compare}
red <- make_cluster_reduction(mask_vol, parcel_map)

spatial_fits <- list(
  parcel_pca = encode(X, spec_space_pca(k = 2), mask = mask_vol,
                      reduction = red),
  hrbf = encode(X, spec_space_hrbf(params = list(
    sigma0 = 2, levels = 0, radius_factor = 2.5, seed = 11L
  )), mask = mask_vol)
)

data.frame(
  family = names(spatial_fits),
  atoms = vapply(spatial_fits, function(x) ncol(loadings(x)), integer(1)),
  rmse = round(vapply(spatial_fits, rmse, numeric(1)), 4)
)
```

```{r validate-spatial, include = FALSE}
spatial_errors <- vapply(spatial_fits, rmse, numeric(1))
stopifnot(all(is.finite(spatial_errors)), all(spatial_errors >= 0))
```

| Spatial structure | Prefer | Why |
|---|---|---|
| Atlas / parcellation is the analysis object | Parcel PCA | Data-driven, per-parcel, block-sparse loadings |
| Shared dictionary reused across subjects | Parcel template | Same atoms for every subject; no per-subject fit |
| Smooth spatial atoms from geometry alone | HRBF | Geometry-driven, no per-parcel model needed |

Use parcel PCA when the atlas or reduction is central to the analysis. Use HRBF
when you want a geometry-driven, smooth spatial dictionary without fitting a
separate PCA model in each parcel.

## How should you choose in practice?

Start from the structure you trust most. If the signal is mostly temporal,
try a temporal basis first. If a shared atlas or spatial dictionary is the
analysis object, start with a spatial basis. If you need both, use
`spec_st()` and check partial reconstruction with `predict()`.

See `vignette("encode-factory")` for the full encoding API and
`vignette("compression-diagnostics")` for storage/error diagnostics.
