---
title: Encoding workflows
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Encoding workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
params:
  family: red

---

```{r setup, include = FALSE}
if (exists("params") && requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(fmrilatent)
```

This vignette walks through the three encoding strategies — temporal,
spatial, and spatiotemporal — and shows how to pick the right one for your
analysis. For a quick introduction, see `vignette("fmrilatent")`.

We use the same small simulated dataset throughout:

```{r shared-data}
mask <- array(TRUE, dim = c(4, 4, 4))
mask_vol <- neuroim2::LogicalNeuroVol(mask, neuroim2::NeuroSpace(dim(mask)))

set.seed(42)
n_time <- 20
n_vox <- sum(mask)
X <- matrix(rnorm(n_time * n_vox), nrow = n_time)
```

## Compressing the time dimension

Temporal specs project each voxel's time series onto a shared set of basis
functions. The data matrix X (time x voxels) becomes a basis B (time x k)
and loadings L (voxels x k), where L = t(X) %*% B. You choose `k` to
control the compression ratio.

### Slepian/DPSS sequences

Discrete prolate spheroidal sequences concentrate energy in a frequency band
while minimizing spectral leakage. Use them for resting-state data or any
analysis focused on a specific frequency range.

```{r slepian-time}
spec_slep <- spec_time_slepian(tr = 2, bandwidth = 0.08, k = 4)
lat_slep <- encode(X, spec_slep, mask = mask_vol, materialize = "matrix")

# 4 basis vectors spanning the 0-0.08 Hz band
dim(basis(lat_slep))

# Reconstruction error
sqrt(mean((as.matrix(lat_slep) - X)^2))
```

The `bandwidth` parameter (in Hz) and `tr` (in seconds) together determine
the normalized half-bandwidth W = bandwidth * tr. The number of
well-concentrated tapers is approximately `2 * n_time * W - 1`; setting `k`
lower than this gives heavier compression.

### DCT

The discrete cosine transform provides orthogonal frequency components. It
is fast and parameter-free beyond choosing `k` — a good default when you
have no strong prior about the signal bandwidth.

```{r dct}
lat_dct <- encode(X, spec_time_dct(k = 8), mask = mask_vol, materialize = "matrix")
dim(basis(lat_dct))
sqrt(mean((as.matrix(lat_dct) - X)^2))
```

### B-splines

B-splines produce smooth, localized basis functions with controllable
flexibility. They are useful for modeling slow drifts or the hemodynamic
response, where you want smooth temporal structure without frequency-domain
assumptions.

```{r bspline-time}
lat_bs <- encode(X, spec_time_bspline(k = 6, degree = 3), mask = mask_vol,
                 materialize = "matrix")
dim(basis(lat_bs))
```

The `degree` parameter controls smoothness (3 = cubic, the default).
Set `orthonormalize = TRUE` (the default) for orthonormal columns, which
simplifies downstream linear algebra.

### Comparing temporal families

All three temporal encodings follow the same pattern. The trade-off is
between spectral precision (Slepian), simplicity (DCT), and localization
(B-spline):

```{r compare-temporal}
rmse <- function(lat) sqrt(mean((as.matrix(lat) - X)^2))

data.frame(
  family = c("Slepian (k=4)", "DCT (k=8)", "B-spline (k=6)"),
  k      = c(ncol(basis(lat_slep)), ncol(basis(lat_dct)), ncol(basis(lat_bs))),
  rmse   = round(c(rmse(lat_slep), rmse(lat_dct), rmse(lat_bs)), 4)
)
```

## Compressing the voxel dimension

Spatial specs reduce the number of spatial locations by projecting voxels
onto a set of spatial atoms. The roles reverse: the loadings L (voxels x k)
define the spatial dictionary, and the temporal coefficients become
B = X %*% L.

### Graph Slepian

Graph Slepian bases are eigenvectors of the graph Laplacian, spectrally
concentrated on spatial clusters. They are the spatial analog of temporal
DPSS sequences.

```{r slepian-space, eval = FALSE}
lat_sp <- encode(X, spec_space_slepian(k = 3, k_neighbors = 6),
                 mask = mask_vol)
```

The `k` parameter sets the number of components per cluster;
`k_neighbors` controls the k-NN graph used for spatial adjacency.

### Heat wavelet

Heat wavelets approximate diffusion on the voxel graph at multiple time
scales, capturing both local and global spatial structure.

```{r heat-space, eval = FALSE}
lat_heat <- encode(X,
  spec_space_heat(scales = c(1, 2, 4, 8), order = 30, threshold = 1e-6),
  mask = mask_vol
)
```

The `scales` parameter controls diffusion time scales (larger = smoother);
`order` sets the Chebyshev polynomial approximation order; `threshold`
truncates small coefficients for sparsity.

### HRBF (hierarchical radial basis functions)

HRBF places smooth Gaussian atoms at data-driven centers. This family gives
you explicit, interpretable spatial basis functions.

```{r hrbf, eval = FALSE}
lat_hrbf <- encode(X,
  spec_space_hrbf(params = list(
    sigma0 = 2, levels = 1,
    radius_factor = 2.5, kernel_type = "gaussian"
  )),
  mask = mask_vol
)
```

The `sigma0` parameter controls atom width; `levels` controls hierarchy
depth (more levels = finer resolution); `radius_factor` sets atom spacing
relative to sigma.

### Wavelet active (CDF 5/3 lifting)

The wavelet active encoder applies CDF 5/3 lifting wavelets along each
spatial axis, touching only in-mask voxels. It is exact (biorthogonal) and
very fast — the best choice for speed-critical pipelines.

```{r wavelet-active, eval = FALSE}
lat_wav <- encode(X,
  spec_space_wavelet_active(levels_space = 2, levels_time = 0, threshold = 0),
  mask = mask_vol
)

# Wavelet active returns an ImplicitLatent; reconstruct with predict()
recon_wav <- predict(lat_wav)
dim(recon_wav)
```

Setting `threshold > 0` zeroes out small wavelet coefficients for lossy
compression. With `threshold = 0` the transform is exact.

## Compressing both dimensions

`spec_st()` combines a temporal and a spatial spec into a separable
representation. Instead of storing the full reconstructed matrix, it stores
a small core tensor and decodes on demand.

```{r separable, eval = FALSE}
lat_st <- encode(X,
  spec_st(
    time  = spec_time_bspline(k = 4, degree = 3),
    space = spec_space_hrbf(params = list(
      sigma0 = 2, levels = 0, radius_factor = 2.5
    ))
  ),
  mask = mask_vol
)

# Full reconstruction
full <- predict(lat_st)
dim(full)

# Partial: only time points 1-5
partial <- predict(lat_st, time_idx = 1:5)
dim(partial)

# Partial: only voxels in an ROI
roi <- array(FALSE, dim = c(4, 4, 4))
roi[1:2, 1:2, 1:2] <- TRUE
roi_data <- predict(lat_st, roi_mask = roi)
dim(roi_data)
```

Partial decoding is the main advantage of separable encoding. You
reconstruct a single time point or a region of interest without paying the
cost of full reconstruction. The core tensor is typically tiny (k_time x
k_space), so storage is minimal regardless of the original data size.

## The convenience factory

`latent_factory()` builds the spec and calls `encode()` in one step. It is
handy for quick experiments:

```{r factory}
lat_f <- latent_factory("dct_time", x = X, mask = mask_vol, k = 8,
                        materialize = "matrix")
dim(basis(lat_f))
```

Available family strings and their underlying specs:

| String | Spec |
|--------|------|
| `"slepian_time"` | `spec_time_slepian()` |
| `"dct_time"` | `spec_time_dct()` |
| `"slepian_space"` | `spec_space_slepian()` |
| `"heat_space"` | `spec_space_heat()` |
| `"slepian_st"` | `spec_st()` with Slepian time + space |
| `"bspline_hrbf_st"` | `spec_st()` with B-spline time + HRBF space |
| `"wavelet_active"` | `spec_space_wavelet_active()` |

Extra arguments in `...` are forwarded to the spec constructor. The two-step
pattern (spec then `encode()`) gives you more control; the factory is a
shortcut when defaults suffice.

## Handle vs. matrix materialization

The `materialize` argument to `encode()` controls how bases are stored
internally:

- **`"handle"`** (default): a lazy object that generates basis columns on
  demand. Saves memory when the basis is large or you only need partial
  reconstruction.
- **`"matrix"`**: an explicit dense matrix. Faster for repeated full
  reconstructions since the basis is precomputed.

```{r materialize, eval = FALSE}
# Eager matrix
lat_eager <- encode(X, spec_time_dct(k = 8), mask = mask_vol,
                    materialize = "matrix")

# Lazy handle (default)
lat_lazy <- encode(X, spec_time_dct(k = 8), mask = mask_vol,
                   materialize = "handle")

# Same numerical result
all.equal(as.matrix(lat_eager), as.matrix(lat_lazy))
#> [1] TRUE
```

Both produce identical numerical results. Choose `"handle"` when memory
matters; choose `"matrix"` when you will call `as.matrix()` or `series()`
repeatedly.

## Benchmarking

`benchmark_roundtrip()` times the encode-decode cycle and reports
reconstruction error across spatial families. It requires the `bench`
package (listed in Suggests).

```{r benchmark, eval = FALSE}
res <- benchmark_roundtrip(
  mask_dims = c(8, 8, 4),
  n_time = 50,
  methods = c("slepian_space", "hrbf", "wavelet_active", "bspline_hrbf_st")
)
res

# Visualize (requires ggplot2)
plot_benchmark_roundtrip(res)
```

The output includes median timing (ms) and reconstruction RMSE for each
family. Use this to find the best trade-off for your data size and accuracy
requirements.

## Quick reference

| Family | Stored | Accuracy | Speed |
|--------|--------|----------|-------|
| Slepian (space) | basis + loadings | exact on span | fast, sparse loadings |
| HRBF | basis + loadings | improves with atom count | moderate, dense atoms |
| Wavelet active | coefficients | exact (biorthogonal 5/3) | very fast, mask-aware |
| Separable (B-spline + HRBF) | core tensor + bases | approximate | fast, separable decode |
| Separable (Slepian) | core tensor + bases | exact on span | fast, sparse spatial |
| Heat wavelet | basis + loadings | approximate (thresholded) | graph build cost, sparse |
