---
title: Getting started with fmrilatent
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fmrilatent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
params:
  family: red

---

```{r setup, include = FALSE}
if (exists("params") && requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(fmrilatent)
```

## Why fmrilatent?

An fMRI dataset is a 4D array — three spatial dimensions and time. Most of
that volume is empty: air, skull, tissue outside the brain. Even within the
brain mask, the signal lives in a much lower-dimensional subspace than the
raw voxel count suggests.

fmrilatent exploits this structure. You choose a basis family, call
`encode()`, and get a `LatentNeuroVec` that stores your data as a compact
matrix factorization. The original data can be reconstructed on demand —
partially or fully — without ever materializing the full dense array.

## A first encoding

Let's compress a small simulated fMRI dataset using a temporal DCT basis:

```{r first-encode}
# Build a small brain mask (4 x 4 x 4, all voxels in-mask)
mask <- array(TRUE, dim = c(4, 4, 4))
mask_vol <- neuroim2::LogicalNeuroVol(mask, neuroim2::NeuroSpace(dim(mask)))

# Simulated data: 20 time points, 64 voxels
set.seed(42)
X <- matrix(rnorm(20 * sum(mask)), nrow = 20)

# Encode with 8 DCT components
lat <- encode(X, spec_time_dct(k = 8), mask = mask_vol, materialize = "matrix")
lat
```

The `LatentNeuroVec` stores a 20 x 8 temporal basis and 64 x 8 spatial
loadings — much smaller than the original 20 x 64 matrix. The savings grow
dramatically with real data: a 300-timepoint run over 100,000 voxels shrinks
from 30 million values to a basis and loadings totaling a fraction of that.

## Getting data back

Reconstruction is straightforward:

```{r reconstruct}
# Full matrix (time x voxels)
recon <- as.matrix(lat)
dim(recon)

# How close is the approximation?
sqrt(mean((recon - X)^2))
```

With 8 out of 20 possible DCT components, the approximation is lossy. More
components capture more variance; fewer give heavier compression.

You can also access data selectively:

```{r access}
# Time series at voxel 1 (linear index)
ts1 <- series(lat, 1L)
head(ts1)

# Inspect the latent factors directly
dim(basis(lat))      # temporal basis: 20 x 8
dim(loadings(lat))   # spatial loadings: 64 x 8
```

## Choosing a basis

Every basis family follows the same two-step pattern: build a spec, then
call `encode()`. The choice of family depends on what structure you want to
preserve.

### Temporal families

Temporal bases compress the time dimension. Each voxel's time course is
projected onto a shared set of basis functions.

```{r slepian-temporal}
# Slepian/DPSS: bandlimited to 0.08 Hz, TR = 2s
lat_slep <- encode(X, spec_time_slepian(tr = 2, bandwidth = 0.08, k = 4),
                   mask = mask_vol, materialize = "matrix")
dim(basis(lat_slep))
```

```{r bspline}
# B-spline: 6 smooth cubic functions
lat_bs <- encode(X, spec_time_bspline(k = 6, degree = 3),
                 mask = mask_vol, materialize = "matrix")
dim(basis(lat_bs))
```

### Spatial and spatiotemporal families

Spatial bases compress the voxel dimension by projecting onto spatial atoms
(graph Slepians, heat wavelets, radial basis functions, or lifting wavelets).
Spatiotemporal bases compress both dimensions into a small core tensor.

```{r spatial, eval = FALSE}
# Spatial Slepian: 3 graph-spectral components per cluster
lat_sp <- encode(X, spec_space_slepian(k = 3, k_neighbors = 6),
                 mask = mask_vol)

# Separable spatiotemporal: B-spline (time) + HRBF (space)
lat_st <- encode(X,
  spec_st(
    time  = spec_time_bspline(k = 4),
    space = spec_space_hrbf(params = list(sigma0 = 2, levels = 0,
                                          radius_factor = 2.5))
  ),
  mask = mask_vol
)

# Partial decode: only time points 1-5
partial <- predict(lat_st, time_idx = 1:5)
dim(partial)
```

Separable encoding returns an `ImplicitLatent` object whose `predict()`
method accepts `time_idx` and `roi_mask` arguments, so you reconstruct only
what you need.

### Parcel-based encoding

When your data comes with a parcellation (Schaefer, Glasser, or any atlas),
you can build a shared spatial dictionary once and project each subject onto
it. The default basis is Laplacian eigenvectors — smooth spatial functions
computed from voxel adjacency within each parcel, with no training data
needed.

```{r parcel-shared, eval = FALSE}
# Define parcels (e.g., from an atlas; here 4 equal parcels)
parcel_ids <- rep(1:4, each = 16)
atlas <- neuroim2::ClusteredNeuroVol(mask_vol, parcel_ids)

# Build a template once — Laplacian eigenvectors, no data needed
tmpl <- parcel_basis_template(atlas, basis_slepian(k = 5))

# Project each subject — same loadings, different scores and offset
lvec_s1 <- encode(X, spec_space_parcel(tmpl), mask = mask_vol)
lvec_s2 <- encode(X, spec_space_parcel(tmpl), mask = mask_vol)  # different data in practice

# Or use data-driven PCA loadings from training data
tmpl_pca <- parcel_basis_template(atlas, basis_pca(k = 5), data = X)
lvec <- encode(X, spec_space_parcel(tmpl_pca), mask = mask_vol)
```

For subject-specific (local) parcel PCA, use `spec_space_pca()` with a
`ClusterReduction`:

```{r parcel-local, eval = FALSE}
red <- as_cluster_reduction(atlas)
lvec <- encode(X, spec_space_pca(k = 5), mask = mask_vol, reduction = red)
```

### Quick reference

| Family | Dimension | Strengths |
|--------|-----------|-----------|
| Slepian/DPSS | time | Optimal bandlimiting, minimal spectral leakage |
| DCT | time | Fast, interpretable frequency components |
| B-spline | time | Smooth, localized, adjustable flexibility |
| Slepian | space | Graph-spectral concentration on regions |
| PCA | space | Data-driven per-parcel components |
| Parcel template | space | Shared dictionary across subjects, atlas-based |
| Heat wavelet | space | Multiscale graph diffusion |
| HRBF | space | Smooth radial basis functions |
| Wavelet active | space | Mask-aware CDF 5/3 lifting, very fast |
| Separable | time + space | Compact core tensor, partial decoding |

## Next steps

- `vignette("encode-factory")` walks through each encoding strategy in
  detail, including benchmarking and materialization options.
- `?LatentNeuroVec` documents the class and its methods (`basis()`,
  `loadings()`, `series()`, `as.matrix()`, `as.array()`).
- `?benchmark_roundtrip` compares encoding families on your own data and
  hardware.
