---
title: Shared Spatial Dictionaries Across Subjects
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
resource_files:
- albers.css
- albers.js
- albers-header.html
params:
  family: red
  preset: homage

vignette: |
  %\VignetteIndexEntry{Shared Spatial Dictionaries Across Subjects} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r library}
library(fmrilatent)
```

Shared spatial dictionaries are useful when multiple subjects should be
represented in the same spatial coordinates. The dictionary is built once from
a template reduction or training sample; each subject gets its own coefficient
time series and offset.

## What is shared?

We start with a nine-voxel mask and three parcels. The training matrix has one
dominant signal in each parcel.

```{r make-template-data}
mask <- array(TRUE, dim = c(3, 3, 1))
mask_vol <- neuroim2::LogicalNeuroVol(mask, neuroim2::NeuroSpace(dim(mask)))
parcel_map <- rep(1:3, each = 3L)
reduction <- make_cluster_reduction(mask_vol, parcel_map)

n_time <- 8L
t <- seq_len(n_time)
signals <- cbind(
  sin(2 * pi * t / n_time),
  cos(2 * pi * t / n_time),
  seq(-1, 1, length.out = n_time)
)
weights <- matrix(c(
  1.0, 0.8, 1.2,
  1.1, 0.9, 1.3,
  0.7, 1.0, 1.4
), nrow = 3, byrow = TRUE)

X_train <- matrix(0, nrow = n_time, ncol = sum(mask))
for (parcel in 1:3) {
  cols <- which(parcel_map == parcel)
  X_train[, cols] <- signals[, parcel] %*% t(weights[parcel, ])
}
```

Build the template once. This example uses one PCA component per parcel so the
render is deterministic and fast.

```{r build-template}
tmpl <- parcel_basis_template(
  reduction,
  basis_pca(k = 1),
  data = X_train,
  center = TRUE
)

validate_template_protocol(tmpl)[c("rank", "n_features", "support_cardinality")]
```

```{r validate-template, include = FALSE}
protocol <- validate_template_protocol(tmpl)
stopifnot(protocol$rank == 3L, protocol$n_features == sum(mask))
```

## How do subjects differ?

Two subject matrices can share the same dictionary while having different
coefficient time series.

```{r encode-subjects}
set.seed(19)
X_subject1 <- X_train + matrix(rnorm(length(X_train), sd = 0.01), nrow = n_time)
X_subject2 <- 0.8 * X_train + matrix(rnorm(length(X_train), sd = 0.01), nrow = n_time)

lat1 <- encode(X_subject1, spec_space_parcel(tmpl), mask = mask_vol)
lat2 <- encode(X_subject2, spec_space_parcel(tmpl), mask = mask_vol)

data.frame(
  subject = c("subject1", "subject2"),
  coefficient_rows = c(nrow(coef_time(lat1)), nrow(coef_time(lat2))),
  coefficient_cols = c(ncol(coef_time(lat1)), ncol(coef_time(lat2))),
  rmse = round(c(
    sqrt(mean((as.matrix(lat1) - X_subject1)^2)),
    sqrt(mean((as.matrix(lat2) - X_subject2)^2))
  ), 6)
)
```

The spatial loadings are the shared part. The coefficients are the
subject-specific part.

```{r shared-loadings}
data.frame(
  same_loading_shape = identical(dim(loadings(lat1)), dim(loadings(lat2))),
  max_loading_difference = max(abs(loadings(lat1) - loadings(lat2))),
  coefficient_difference = max(abs(coef_time(lat1) - coef_time(lat2)))
)
```

```{r validate-subjects, include = FALSE}
stopifnot(
  max(abs(loadings(lat1) - loadings(lat2))) < 1e-10,
  max(abs(coef_time(lat1) - coef_time(lat2))) > 0
)
```

Use `vignette("shared-structure-boldzip")` when you need the lower-level
shared-structure contracts and archive handoff boundaries.
