---
title: Standalone Codecs
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
vignette: |
  %\VignetteIndexEntry{Standalone Codecs} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(fmrilatent)
```

The primary package workflow is `spec_*()` plus `encode()`, which returns
latent neuroimaging representations such as `LatentNeuroVec` and
transport-backed latents.

Standalone codecs are different. They work on ordinary matrices, own their
own encode/decode payloads, and are not dispatched through `encode()`.
BOLDZip-SR is currently the only standalone codec in `fmrilatent`.
For a full encode/decode/evaluate workflow, see `vignette("boldzip")`.

## What sets a standalone codec apart?

BOLDZip-SR is currently the only standalone codec in `fmrilatent`, and it
makes the category concrete. Three properties distinguish it from a
`spec_*()` encoder:

- **Matrix in, matrix out.** It expects voxels (or grayordinates) in rows and
  time in columns — intentionally the transpose of the `encode()` convention,
  which uses time by voxels.
- **Owns its payload.** The encoder returns a `BoldZipSR` object, not a
  `LatentNeuroVec`, and it carries its own decode logic.
- **Not in the encode-spec registry.** You call it directly; it is never
  dispatched through `encode()`.

A minimal fit shows all three at once:

```{r boldzip-small}
set.seed(9)
n_voxels <- 10L
n_time <- 24L
carrier <- sin(seq(0, 2 * pi, length.out = n_time))
texture <- seq(-1, 1, length.out = n_voxels)
X <- outer(texture, carrier)               # voxels x time
X <- X + matrix(rnorm(n_voxels * n_time, sd = 0.01), nrow = n_voxels)

fit <- boldzip_sr_encode(X, k_carriers = 1L, q_texture = 1L,
                         temporal_k = 6L)

inherits(fit, "BoldZipSR")    # a codec object, not a LatentNeuroVec
inherits(fit, "LatentNeuroVec")
```

That is the whole distinction. For the full encode → evaluate → decode →
budget-tuning workflow (including the `as_implicit_latent()` bridge back into
the package's decoder contract), see `vignette("boldzip")`. For the
lower-level shared-structure contracts BOLDZip-SR can share with other
representations, see `vignette("shared-structure-boldzip")`.
