---
title: Transport-Aware Encoding
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
resource_files:
- albers.css
- albers.js
- albers-header.html
params:
  family: red
  preset: homage

vignette: |
  %\VignetteIndexEntry{Transport-Aware Encoding} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r library}
library(fmrilatent)
```

Transport-aware encoding separates the shared basis from the subject-specific
field operator. `fmrilatent` owns coefficient recovery and decode/projection;
the operator that maps template fields into the subject target domain is
compiled outside the package.

## What basis are we transporting?

This small example uses six voxels split into two parcels. A one-component
PCA dictionary per parcel gives a two-column shared basis asset.

```{r make-basis}
mask <- array(TRUE, dim = c(3, 2, 1))
mask_vol <- neuroim2::LogicalNeuroVol(mask, neuroim2::NeuroSpace(dim(mask)))
parcel_map <- rep(1:2, each = 3L)
reduction <- make_cluster_reduction(mask_vol, parcel_map)

X_template <- matrix(c(
  1, 2, 2, 1, 0, 0,
  0, 1, 1, 0, 1, 1,
  1, 0, 0, 1, 2, 2,
  0, 2, 2, 0, 1, 1,
  1, 1, 0, 0, 2, 1,
  0, 0, 1, 1, 1, 2
), nrow = 6, byrow = TRUE)

tmpl <- parcel_basis_template(reduction, basis_pca(k = 1),
                              data = X_template, center = TRUE)
```

## What does the field operator look like?

For a real study, this object would come from registration and sampling code.
Here the template and subject grids match, so the operator is identity.

```{r identity-operator}
identity_operator <- list(
  n_source = sum(mask),
  n_target = sum(mask),
  source_domain_id = "template:demo",
  target_domain_id = "subject01:bold",
  forward = function(z) as.matrix(z),
  adjoint_apply = function(y) as.matrix(y),
  provenance = list(operator = "identity-demo", target_mask = mask_vol)
)
```

## How are coefficients recovered?

`encode_transport()` solves for coefficient time series in the shared analysis
coordinates and carries run metadata forward.

```{r encode-transport}
run_info <- list(run_id = rep(c(1L, 2L), each = 3L))

lat_tx <- encode_transport(
  X_template,
  basis_asset = tmpl,
  field_operator = identity_operator,
  lambda = 0,
  run_info = run_info
)

data.frame(
  time_points = nrow(coef_time(lat_tx, "analysis")),
  analysis_components = ncol(coef_time(lat_tx, "analysis")),
  finite = all(is.finite(coef_time(lat_tx, "analysis")))
)
```

```{r validate-transport, include = FALSE}
stopifnot(
  ncol(coef_time(lat_tx, "analysis")) == template_rank(tmpl),
  all(is.finite(coef_time(lat_tx, "analysis")))
)
```

## What is handed downstream?

Downstream model-fitting code should consume analysis coordinates. When a
model returns a coefficient-space effect, `decode_coefficients()` projects it
back to native or template space.

```{r decode-effect}
gamma <- coef_time(lat_tx, "analysis")[1, ]
native_map <- decode_coefficients(lat_tx, gamma, space = "native", wrap = "none")

data.frame(
  coefficient_length = length(gamma),
  native_values = length(native_map),
  finite_native = all(is.finite(native_map))
)
```

This is the package boundary: external code builds the field operator and fits
the statistical model; `fmrilatent` keeps the shared basis, coefficients, and
decode semantics coherent.
